---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data

Load data and create the dataframes used along the document. The only transformation is a cast to *Date* type for column *date*: 

```{r carica_dati}
knitr::opts_chunk$set(fig.path = "figure/") # save directory for plots

activity = read.csv("activity.csv")
activity$date <- as.Date(activity$date) # cast to Date

stepsPerDay = with(activity[!is.na(activity$steps), ], aggregate(steps, list(date), sum))
names(stepsPerDay) <- c("date", "steps.per.day")

stepsPerInterval = with(activity[!is.na(activity$steps), ], aggregate(steps, list(interval), mean))
names(stepsPerInterval) <- c("interval", "mean.steps")

meanStepsPerDay = with(activity[!is.na(activity$steps), ], aggregate(steps, list(date), mean))
names(meanStepsPerDay) <- c("date", "mean.per.day")

meanStepsPerInterval = with(activity[!is.na(activity$steps), ], aggregate(steps, list(interval), mean))
names(meanStepsPerInterval) <- c("interval", "mean.per.day")

```

**stepsPerDay** sum of all steps in the day per day. NA values are omitted

**stepsPerInterval**/**meanStepsPerInterval**: mean of steps of all days per time interval NA values are omitted

**meanStepsPerInterval**: mean of steps of all days per time interval NA values are omitted

## What is mean total number of steps taken per day?
We can calculate mean and median number of steps taken each day:
```{r calcola_medie}
# calculate median and mean
mystat <- summary(stepsPerDay$steps.per.day)
mdn <- mystat["Median"]
mn <- mystat["Mean"]
```
Mean of steps per day: **`r sprintf("%.2f",mn)`**

Median of steps per day: **`r sprintf("%.2f", mdn)`**

# plot histogram of total number steps per day
Using the *stepsPerDay* defined above we can plot total number steps per day

```{r hist_noNA}

hist(stepsPerDay$steps.per.day, breaks = 20, col = "yellow", xlab = "", main = "Total number of steps taken each day (NA excluded)")

```



## What is the average daily activity pattern?

We can plot the average steps per interval:

```{r average_steps_per_day}
with(stepsPerInterval, plot(interval, mean.steps, type = "l", xlab = "Time interval (5 min)", ylab = "Average steps per day", main = "Average steps taken per day @5 minutes intervals"))

# calculate the maximun number of steps in the interval
maxstepsidx <- which.max(stepsPerInterval$mean.steps)
maxsteps <- stepsPerInterval$mean.steps[maxstepsidx]

```
and calculate the maximun number of steps in the interval: **`r sprintf("%.2f", maxsteps)`**

## Imputing missing values
First we calculate the number of missing values (NA):
```{r total_number_NA}

# Number of rows with NA
totna <- with(activity, sum(is.na(steps)))

```
Number of rows with **NA**: `r totna`

Then we can impute the NA in a given interval using the mean value for that 5-minutes interval  calculate across all days. 

```{r average_withNA}

# Create a new dataset that is equal to the original dataset but with the missing data filled in
mact <- merge(activity, meanStepsPerInterval, by = "interval")
mact[is.na(mact$steps), 2] <- mact[is.na(mact$steps), 4]
# Calculate steps per day
newStepsPerDay = aggregate(mact$steps, list(mact$date), sum)
names(newStepsPerDay) <- c("date", "steps.per.day")
# Calculate Mean and Median
mynastat <- summary(newStepsPerDay$steps.per.day)
mdnna <- mynastat["Median"]
mnna <- mynastat["Mean"]
```
Mean of steps per day (NAs imputed): **`r sprintf("%.2f", mnna)`** - $Mean_{NAimputed} - Mean_{NAexcluded} =  `r mnna - mn`$

Median of steps per day (NAs imputed): **`r sprintf("%.2f", mdnna)`** - $Median_{NAimputed} - Median_{NAexcluded} =  `r mdnna - mdn`$

Finally we can plot the same histogram as above but with NA imputed
```{r hist_withNA}

# same histogram as above but with new dataset
hist(newStepsPerDay$steps.per.day, 20, col = "yellow", xlab = "", main = "Total number of steps taken each day (NA imputed)")

```



With NA imputed using mean values in the 5-minutes interval across all days Median increased compared to that with NA excluded.  


## Are there differences in activity patterns between weekdays and weekends?

To check differences in activity patterns between weekdays and weekends we can create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r factor_week}
# Set locale to "English". Only for NON English LOCALE
loc <- Sys.setlocale("LC_ALL", "en_US.UTF-8") # Linux
#loc <- Sys.setlocale("LC_ALL", "English") # Windows

# create dataset 
ac1 = mact[ , -c(4)]
ac1$wd <- weekdays(mact$date)

# prepare data for factor variable: 1 = weekday, 2 = weekend
ac1[ac1$wd %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),4] <- 1
ac1[ac1$wd %in% c("Saturday", "Sunday"),4] <- 2

# create new factor variable WeekDay/WeekEnd
ac1$wk <- factor(ac1$wd, levels = c(1,2), labels = c("WeekDay", "WeekEnd"))

```

and now we can calculate the means ad plot WeekDay vs WeekEnd
```{r last_plot}

# Calculate mean
myplotweek <- aggregate(steps ~ interval + wk, data=ac1, mean)
require(ggplot2)
qplot(interval, steps, data = myplotweek, facets = wk ~ ., geom = "line")

```

